Staged Abstract Interpreters

### 2019-03-07 paper/10_intro.tex
update

Now we have seen two orthogonal transformations of interpreters: by
specialization, we can refactor interpreters to code generators; by approximation,
we can refactor interpreters to static analyzers. As an intellectual quest, it
is natural to ask --- starting from an interpreter, \textit{can we derive a
sound abstract interpreter while it produces fast code that does the analysis?}
Conceptually, it is similar to the concrete setting: for an
abstract interpreter $\widehat{\texttt{eval}}: \texttt{Expr} \to
\widehat{\texttt{Env}} \to \widehat{\texttt{Value}}$, which takes a program, an abstract
environment and returns abstract values, we would like to specialize it with
respect to a program $e_0$ and produce a compiled analysis
$\widehat{\texttt{eval}}_{\texttt{e0}} : \widehat{\texttt{Env}} \to
\widehat{\texttt{Value}}$, such that $
\widehat{\texttt{eval}}_{\texttt{e0}}(\widehat{\rho}) =
\widehat{\texttt{eval}}(e_0)(\widehat{\rho})$, but the compiled one leaves the
abstract environment $\widehat{\rho}$ as remaining input.

@@

In this paper, we study the confluence of two old ideas --- Futamura projection
and abstract interpretation, but from the perspective of their recent
realizations --- multi-stage programming and definitional abstract interpreters.
To be specific, we present the application of the first Futamura projection on
definitional abstract interpreters, which intersects as staged abstract
interpreters. By borrowing the ideas from monadic interpreters
\cite{Steele:1994:BIC:174675.178068, DBLP:conf/popl/LiangHJ95,
DBLP:journals/pacmpl/DaraisLNH17, Sergey:2013:MAI:2491956.2491979} and embedded
domain-specific languages \cite{DBLP:conf/snapl/RompfBLSJAOSKDK15,
DBLP:journals/jfp/CaretteKS09, DBLP:conf/icfp/GibbonsW14, Hofer:2008:PED:1449913.1449935}, we develop an
approach to fulfill four different semantics (Figure~\ref{confluence}) in a
unified framework. The four semantics share one same generic interpreter.
However, by instantiating the interpreter differently, it can behave as a
concrete interpreter, an abstracter interpreter, or their code-generator
versions, respectively. The key idea that enables this flexibility is to use
abstract type members abstract over concrete/abstract components (e.g., concrete
values or abstract values), as well as the binding-time of them (e.g., static or
dynamic). By staging, the overhead caused by the monadic layers is eliminated in
the generated code.

\paragraph{Performant via Specialization}

Besides the intellectual merit shown in the confluence diagram, this paper
contributes to a practical problem in static analysis: \textit{can we speed up a
static analyzer without any intrusive changes and soundness compromises?} Static
analysis is known as a trade-off game between \textit{precision} and
\textit{efficiency}, we argue that applying meta-programming (multi-stage
programming, in particular) with abstract interpreters is an effective approach
to improve the latter one while the former one is untouched. Specialization
removes the overhead of abstractions introduced by high-level programming, but
not changes the meaning of the program -- in this case, the abstract semantics
implemented by the analyzer. Indeed, specializing analysis with respect to a
program by partial evaluation \cite{damian1999partial, amtoft1999partial,
Boucher:1996:ACN:647473.727587, ashley:practical} is not a new idea, despite the
fact that most of the previous work are ad-hoc on a particular analysis or
requiring significant changes to the analysis. For example, abstract compilation
\cite{Boucher:1996:ACN:647473.727587} requires the whole analyzer to be
rewritten as closure generation from. In this paper, we show a systematic
approach to meta-programming achieves the performance goal with less intrusive
changes; in fact, a case study (Section~\ref{cs_ac}) comparing with abstract
compilation shows, by utilizing type-based stage annotations, the analyzer
program does not need any changes.

@@

\todo{modular analysis: a whole-program abstract
  interpreter can be run on different components of program separately, we
  achieve this by meta-programming mechnically}

@@

When staging a concrete interpreter, the programmers need to distinguish static
and dynamic values --- the given program to be executed by the interpreter is
classified as static because it is known at compile-time, and the inputs to that
program are dynamic. However, when staging an abstract interpreter, this
distinction does not exist anymore. Because the abstract interpreter
instantiates all the inputs as some form of abstract values, which are usually
top elements in their abstract domains and are also statically known. Then what
is the point of staging if there is no such distinction? A surprising by-product
of thinking about this question is to realize that we can apply the staged
abstract interpreter on \textit{open} programs, and the free variables
representing other parts of the program (e.g., libraries) are dynamic inputs,
therefore we obtain a modular analysis through staging, mechanically.
\note{TR: I don't understand this. The program structure is static, the abstract values
are still dynamic, no? They change in every iteration of the fixpoint algorithm}
\note{TR: why is it a big deal. On Closed programs we obtain a constant factor,
but modular analysis has different asymptotics}
\fi

@@

\iffalse
It has been observed that partially applying context-sensitivity on selected
portion of the program could improve the precision and efficiency
\cite{zipper2018, Kastrinis:2013:HCP:2491956.2462191}. We show that staging
abstract interpreters as an approach to effectively implement hybird
context-sensitivity\todo{}.
\fi

\note{
  Expression problem \cite{DBLP:conf/icfp/GibbonsW14, DBLP:conf/ecoop/KrishnamurthiFF98, expproblem},
  deep embedding, four different interpretations.
}

>>>>

Now we have seen two orthogonal transformations of interpreters: by
specialization, we can refactor interpreters to code generators; by
approximation, we can refactor interpreters to static analyzers. As an
intellectual quest, we should naturally ask -- starting from an interpreter,
\textit{can we derive a sound abstract interpreter while it produces fast code
that does the analysis?}
Conceptually, it is similar to the concrete setting: for an
abstract interpreter $\widehat{\texttt{eval}}: \texttt{Expr} \to
\widehat{\texttt{Env}} \to \widehat{\texttt{Value}}$, which takes a program, an abstract
environment and returns abstract values, we would like to specialize it with
respect to a program $e_0$ and produce a compiled analysis
$\widehat{\texttt{eval}}_{\texttt{e0}} : \widehat{\texttt{Env}} \to
\widehat{\texttt{Value}}$, such that $
\widehat{\texttt{eval}}_{\texttt{e0}}(\widehat{\rho}) =
\widehat{\texttt{eval}}(e_0)(\widehat{\rho})$, but the compiled one leaves the
abstract environment $\widehat{\rho}$ as remaining input.

@@

In this paper, we study the confluence of two old ideas --- Futamura projection
and abstract interpretation, but from the perspective of their recent
realizations --- multi-stage programming and definitional abstract interpreters.
To be specific, we present the application of the first Futamura projection on
definitional abstract interpreters, which intersects as staged abstract
interpreters. By borrowing the ideas from monadic interpreters
\cite{Steele:1994:BIC:174675.178068, DBLP:conf/popl/LiangHJ95,
DBLP:journals/pacmpl/DaraisLNH17, Sergey:2013:MAI:2491956.2491979} and embedded
domain-specific languages \cite{DBLP:conf/snapl/RompfBLSJAOSKDK15,
DBLP:journals/jfp/CaretteKS09, DBLP:conf/icfp/GibbonsW14,
Hofer:2008:PED:1449913.1449935}, we develop an approach to fulfill four
different semantics (Figure~\ref{confluence}) in a unified framework. The four
semantics share one same generic interpreter interface. Later, by instantiating
the interpreter differently, it can behave as a concrete interpreter, an
abstracter interpreter, or their code-generator versions, respectively. The key
idea that enables this flexibility is to use abstract type members abstract over
concrete/abstract components (e.g., concrete values or abstract values), as well
as the binding-time of them (e.g., static or dynamic). By staging, the overhead
caused by the monadic layers is eliminated in the generated code.

\paragraph{Performance via Specialization}

Besides the intellectual merit shown in the confluence diagram, this paper
contributes to a practical problem in static analysis: \textit{can we speed up a
static analyzer without any intrusive changes and soundness compromises?} Static
analysis is known as a trade-off game between \textit{precision} and
\textit{efficiency}, we argue that applying meta-programming (multi-stage
programming, in particular) with abstract interpreters is an effective approach
to improve the latter one while the former one is untouched. Specialization
removes the overhead of abstractions introduced by high-level programming (e.g.,
monads), but not loses any advantages of this programming style (e.g.,
extensibility), nor changes the meaning of the program -- in this case, the
abstract semantics that does the analysis. Indeed, specializing analysis
with respect to a program by partial evaluation \cite{damian1999partial,
amtoft1999partial, Boucher:1996:ACN:647473.727587, ashley:practical} is not a
new idea, despite the fact that most of the previous work are ad-hoc on a
particular analysis or requiring significant changes to the analysis. For
example, abstract compilation \cite{Boucher:1996:ACN:647473.727587} requires the
whole analyzer to be rewritten as closure generation from. In this paper, we
show a systematic approach to meta-programming that achieves the performance
goal with less intrusive changes; in fact, a case study (Section~\ref{cs_ac})
comparing with abstract compilation shows, by utilizing type-based stage
annotations, the analyzer program does not need any changes.

@@

Our second contribution is to enable modular analysis based on a whole-program
analyzer by meta-programming. Modern software are shipped with large libraries
\cite{toman_et_al:LIPIcs:2017:7121}, and analyzing programs with large libraries
is usually expensive or unnecessarily repeated for whole-program analyzing
algorithms. For example, it has been shown that analyzing a simple ``Hello
World'' program in Java depends on additional 3,000 classes in the library
\cite{DBLP:conf/oopsla/KulkarniMZN16}. Meta-programming gives us a chance to
tweak a whole-program analyzing algorithm into an analyzer that can work on
different components of program separately. Still, the idea is to specialize the
abstract interpreter used for whole-program analysis, but on a subset of the
program, for examples, several generic functions for list manipulation. Such
specialization works on a lambda expression and generates code, which is partial
analysis results and can be completed by running with the necessary arguments
(for example, an environment that closes the lambda term) at next stages. Multiple
partial analysis results also can be reused later and composed together as they
are just normal programs at the next stage. Through staging, we mechanically
obtain a modular analysis, even though the original algorithm is formulated as a
whole-program analysis.

@@

When staging a concrete interpreter, the programmers need to distinguish static
and dynamic values --- the given program to be executed by the interpreter is
classified as static because it is known at compile-time, and the inputs to that
program are dynamic. However, when staging an abstract interpreter, this
distinction does not exist anymore. Because the abstract interpreter
instantiates all the inputs as some form of abstract values, which are usually
top elements in their abstract domains and are also statically known. Then what
is the point of staging if there is no such distinction? A surprising by-product
of thinking about this question is to realize that we can apply the staged
abstract interpreter on \textit{open} programs, and the free variables
representing other parts of the program (e.g., libraries) are dynamic inputs,
therefore we obtain a modular analysis through staging, mechanically.
\note{TR: I don't understand this. The program structure is static, the abstract values
are still dynamic, no? They change in every iteration of the fixpoint algorithm}
\note{TR: why is it a big deal. On Closed programs we obtain a constant factor,
but modular analysis has different asymptotics}

@@

It has been observed that partially applying context-sensitivity on selected
portion of the program could improve the precision and efficiency
\cite{zipper2018, Kastrinis:2013:HCP:2491956.2462191}. We show that staging
abstract interpreters as an approach to effectively implement hybird
context-sensitivity\todo{}.
\fi

\note{
  Expression problem \cite{DBLP:conf/icfp/GibbonsW14, DBLP:conf/ecoop/KrishnamurthiFF98, expproblem},
}

### 2019-03-07 paper/21_bg_interp.tex
update

In Scala, thanks for its nature that combines functional programming and
object-oriented programming, we may use traits and objects to implement monads.
A trait poses monadic operations as abstract methods that objects extending it
should implement. An object encapsulates the data and implements monadic methods
that operate on these data. The monadic @bind@ operation is usually called
@flatMap@ in Scala. For example, the following trait @Monad@ declares the
@return@ and @flatMap@ operation:

\begin{lstlisting}
  trait Monad[F[_]] {
    def return[A](a: => A): F[A]
    def flatMap[A,B](fa: F[A])(f: A => F[B]): F[B]
  }
\end{lstlisting}

An object may encapsulate the data of type @F[A]@ and only provide
@flatMap(f: A => F[B]): F[B]@. This allows us to utilize the for-comprehensions in
Scala, which provide special syntactic support for moandic @flatMap@.
For example, a list object of element type @A@ is a monad and it implements the
methods @flatMap[B](f: A => List[B]): List[B]@ and @map[B](f: A => B): List[B]@.
To compute the Cartesian product of two lists of numbers, we may directly call
@flatMap@ and pass an anonymous function that internally calls the monadic
@return@ operation (the @List@ constructor in this case):

@@

In the rest of the paper, we will use for-comprehension syntax for monadic
operations, the implementation of monads and monad transformers essentially
borrows the ground-truth from Haskell.

>>>>

In Scala, thanks for its nature that combines functional programming and
object-oriented programming, we may use traits and objects to implement monads.
For example, the following trait @Monad@ declares the @pure@ and @flatMap@
operation. The trait itself takes a higher-kinded type @F[_]: * -> *@
which takes a type and returns a type. The method @pure@ promotes values of type
@A@ to values of type @F[A]@. The monadic @bind@ operation is usually called
@flatMap@ in Scala, which takes a monad-encapsulated value of type @F[A]@, a function of
@A => F[B]@ and returns values of type @F[B]@.

\begin{lstlisting}
  trait Monad[F[_]] {
    def pure[A](a: A): F[A]
    def flatMap[A,B](fa: F[A])(f: A => F[B]): F[B]
  }
\end{lstlisting}

An object implementing the trait @Monad@ may also encapsulates the data of type
@F[A]@ internally, in this case, it only needs to provide method
@flatMap(f: A => F[B]): F[B]@. This allows us to utilize the for-comprehensions
in Scala, simliar to Haskell's do-notation, which provides special syntactic
support for moandic @flatMap@. For example, a list object of element type @A@ is
a monad and it implements the methods @flatMap[B](f: A => List[B]): List[B]@ and
@map[B](f: A => B): List[B]@. To compute the Cartesian product of two lists of
numbers, we may directly call @flatMap@ and pass an anonymous function that
internally calls the monadic @pure@ operation (the @List@ constructor in this
case):

@@

In the rest of the paper, we use Scala's for-comprehension syntax and monad
transformers @ReaderT@, @StateT@, @ListT@ to write our interpreters. The
implementation of monads and monad transformers essentially borrows the
ground-truth from Haskell.
